---
title: "Estimating Species Composition"
author: "Paul van Dam-Bates"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: >
  An overview of how to estimate species composition using `SonarSpeciesComp`.
vignette: >
  %\VignetteIndexEntry{Using nimbleQuad}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo= FALSE,
  comment = "#>",
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "70%",
  fig.align = "center"
)
```

## Introduction

The R package `SonarSpeciesComp` is intended to estimate species composition for fish observed by the hydroacoustic monitoring programs at either Qualark (Fisheries and Oceans Canada), or Mission (Pacific Salmon Commission)

The basic idea is to use hydroacoustic length measurements combined with nearby test fishery counts to identify the proportion of each species migrating each day. A day at Mission runs from 5:00 - 4:59 and at Qualark it is 00:00 - 23:59. 


## Data Types

### Sonar Counts

Every hour hydroacoustic counts are taken in each sonar bin using an ARIS 1800. At Qualark the sonar is rolled 35 degrees to be able to enumerate all fish along a steep bank. At Mission, the right bank is relatively flat and a single aim with no roll is sufficient for full coverage. However, on the left bank, two aims are used within the first 10 m to enumerate all fish. At both sites, counts are done by bank, and 10 m distance bins. Qualark completes counts out to 30 m, or 3 bins on each bank, and Mission out to 50 m. Counts are labelled as Up and Down for the number of fish counted as swimming upstream or down stream.

The package requires count data as a data frame with columns that are slightly specific for each site.

1. Mission requirements:

```{r}
df <- data.frame(Date = as.Date("2023-09-01", format = "%Y-%m-%d"), MissionDate = as.Date("2023-09-02", format = "%Y-%m-%d"), Hour = 5, SonarCode = "A1", SonarAim = "Aim1", SonarBin = "Bin1", Bank = "Left Bank", Up = 5, Down = 3, Duration = 5)
df
```

2. Qualark requirements:

```{r}
df <- data.frame(Date = as.Date("2023-09-01", format = "%Y-%m-%d"), Hour = 5, SonarBin = "Bin1", Bank = "Left Bank", Up = 5, Down = 3, Duration = 20)
df
```

### Sonar Lengths

1. Mission requirements:

```{r}
df <- data.frame(Date = as.Date("2023-09-01", format = "%Y-%m-%d"), MissionDate = as.Date("2023-09-02", format = "%Y-%m-%d"), Hour = 5, SonarCode = "A1", SonarAim = "Aim1", SonarBin = "Bin1", Bank = "Left Bank", R.m = 6.3, L.cm = 55.1)
df
```

2. Qualark requirements:

```{r}
df <- data.frame(Date = as.Date("2023-09-01", format = "%Y-%m-%d"), Hour = 5, SonarBin = "Bin1", Bank = "Left Bank", R.m = 6.3, L.cm = 55.1)
df
```

## Predicting species proportions:

List of model matrix for each species. Column of 1 if constant. Many might change by near shore, off shore etc.

