---
title: "Species Composition Model Workflow at Mission"
description: >
  Workflow for estimating species composition at Mission.
format: 
  html:
    embed-resources: true
    theme: cosmo
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Species Composition Model Workflow at Mission}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r include = FALSE}
library(SonarSpeciesComp)
```

## Introduction

The R package `SonarSpeciesComp` is designed to estimate species composition, the number of each salmon species migrating past Mission, or Qualark, daily. The package can be installed using either `remotes` or `devtools` directly from GitHub as:

```{r, echo = TRUE, eval = FALSE}
## Build vignette if you want to:
remotes::install_github("paul-vdb/SonarSpeciesComp")
library(SonarSpeciesComp)
```

The package relies on building an `R6` class object, storing all the necessary information within that object, and then running methods within to fit the joint species composition model. By joint, we refer to jointly modeling test fishery catch with sonar length measurements. See `?speciesCompModel` for all the available field and methods within the model object.

To start, build the basic model object as:
```{r}
obj <- speciesCompModel$new(date = "2022-08-01", site = "Mission")
```
This object will default to:

1. Species: `r obj$species_info$species`.
1. Test fishery parameters `r obj$data_info$test_fishery_input`, which results from the default "formula" varies the test fishery parameters by `r obj$data_info$test_fishery_formula`.
1. Roll angle: `r obj$data_info$roll_angle` degrees.
1. Feasible lengths: `r obj$data_info$feasible_lengths` cm.
1. Date to estimate: `r obj$est_date` is what was set, but it defaults to today's date but a value should be provided.

## Data Input

At Mission, there are standardized input based on daily workflow that the model relies on. These are:

1. `MergedLengths.csv` - The measured sonar lengths for the season up until the current day.
1. `SONARMERGE export.csv` - The sonar counts for the season up until the current day.
1. `WhonnockSpeciesComp.xlsx`, sheet = "by_set" - Test fishery catch which has date as "TRIP_DTT", set number as "FE_SET_NO" and includes species by gilled, tangled as well as "net_length", "start_out", "full_out", "start_in", and "full_in".
1. `Whonnock Fork Lengths OtherSalmon.xlsx` - Fork lengths for Chinook (cm).
1. `2025ABfiles.xlsx` - Excel file with sockeye and pink fork lengths (cm). We expect the column "Species" to be c("Sockeye" and "Pink"). These two length files should be joined with a new column "FL.cm" to be the fork length in cm and contains a "Date" column.
1. `Chinook and Pinks.xlsx`, sheet = "2. Salmon Passage by Bin" - Salmon passage table with columns "MissionDate" as a date and a computed column "total" that is a sum over both bins and offshore (mobile).
1. `Albion Chinook catch and set mesh 2016-2025.xlsx` - Albion test fishery data with columns "CRPT_DTT", "NET_START_OUT", "NET_FULL_OUT", "NET_START_IN", "NET_FULL_IN", SET_NO", "NET_CONFIG", and "CATCH_QTY".

The package has two years of data pre-loaded as example of what format the data should look like at Mission to successfully run the model.

```{r}
data(mission_2022, package = "SonarSpeciesComp")
data(mission_2025, package = "SonarSpeciesComp")
```

Let's now add data to the model object via `$processData`

```{r}
obj$processData(sonar_counts = mission_2022$sonarCounts, 
  sonar_lengths = mission_2022$sonarLengths,
  test_fishery_counts = list(whonnock = mission_2022$whonnock, albion = mission_2022$albion), 
  salmon_passage_table = mission_2022$salmonPassage, 
  include_tangled = TRUE, dropN = 2, feasible_lengths = c(10,120))
```

Note that here we choose to include fish that are tangle, not just gilled by the test fishery. We drop lengths if only 2 were measured, which made a difference previously in some models but should not be necessary. Feasible lengths can be chosen during processing but is also held as internally on setup.

We have now built a model object that contains the entire data set and is ready to run for individual days under varying parameterizations, including changing the number of species.

## Running the Model

### Setting the data:

We will now choose a specific date, and the species present. In 2022 at Mission small resident fish were included. Note that species are always added in lower case with no spaces (e.g. "smallresidentfish"). We should also choose the number of days to combine in analysis to help fit the expansion line estimates.

```{r}
obj$setDate(date = "2022-08-01", ndays = 3)
obj$setSpecies(species = c("smallresident", "largeresident", "jackchinook", "sockeye", "adultchinook"))
```
The next step is to choose the species mean and standard deviation lengths that will be held fixed when fitting the mixture model. Standard values are held as default in the model as:

```{r}
default_params()$mu
default_params()$sigma
```

The mean (`mu`) and standard deviation (`sigma`) values are then set as:
```{r}
obj$setSpeciesLengths(mu = c("smallresident" = 22, "largeresident" = 38, "jackchinook" = 42), 
  sigma = c("smallresident" = 2.5, "largeresident" = 3.5), 
  test_fishery_lengths = mission_2022$testFisheryLengths, ndays = 7)
```

Note that within this function we fit a mixture model to the Chinook catch from the test fishery to see if we can set a reasonable proportion of jack Chinook vs adult (or small and large) Chinook.

### Parameter Setup

The user is able to influence how the species composition model is set up mostly through the `setModelParameters` function. The input includes:

1. `fixed_parameters` - Here we can choose which parameters to hold completely fixed. We suggest to always choose `c("mu", "sigma")`and current default includes "proportion_jackchinook". Note that holding proportion jack Chinook fixed, still allows the proportion of total Chinook to vary, meaning it doesn't fix the total number of jack Chinook, just the proportion relative to adult Chinook. Note that there will always be `$proportion_adultchinook` that is held fixed at one if only adult Chinook are present in the model or by the user (potentially informed by the test fishery) when including small and large adult Chinook.

```{r}
obj$setModelParameters(fixed_parameters = c("mu", "sigma"))
obj$params_fixed
```

2. `fixed_values` - List with names equal to parameters that the user can use to set the actual values to fix. Otherwise, the parameters chosen in `fixed_parameters` will be chosen based on current values held within the model object. Additional individual values in parameters can be held constant here. For example, we could choose to change the small resident size here, and potentially fix measurement error $\sigma_0 = 4$. Note that the term will appear transformed for parameters that are fit on the real scale.

```{r}
obj$setModelParameters(fixed_parameters = c("mu", "sigma"),
  fixed_values = list(sigma0 = 4))
obj$params_fixed
```

3. `initial_values` - List of what to set the initial values to for the model. Things that are not set will be given default values but it may sometime help to initialize the model to make fitting faster. The default beam spreading formula is `beta[1] + beta[2]*beamWidth.cm`, and we may want to set `beta = c(0,0.8)`.

```{r}
obj$setModelParameters(fixed_parameters = c("mu", "sigma"), 
  fixed_values = list(), 
  initial_values = list(beta = c(0, 0.8), sigma0 = 4))
obj$params_init$log_sigma0
obj$params_init$beta
```

4. `formula_proportions` - List of formulas to fit the proportions of each species to covariates. Choose `list("all" = ~ 0)` to set all proportions constant, excluding Chinook. If multiple days then recommend `"all" ~ 0 + factor(day)`, however we may choose to assume the proportion of jack Chinook remains constant `"jackchinook" = ~ 1`.

```{r}
obj$setModelParameters(fixed_parameters = c("mu", "sigma"), 
  initial_values = list(beta = c(0, 0.8), sigma0 = 4),
  formula_proportions = list("all" = ~ 0 + factor(day), "jackchinook" = ~ 1))
obj$data_list$alpha_formula
```

5. `date` and `ndays` - Option to be able to update the day for analysis here. Just another way to call `setDate`.

6. `species` - Option to change the species here as an alternative to `setSpecies`. 

7. `delta_mu_bounds` - List of formulas to fit the proportions of each species to covariates. Choose `~ 0 + SonarBin` to let it vary by each Sonar Bin.

8. `qinv_names` - Choose how the expansion line should vary by species, fishery, and net_type. If we think that all Albion test fishery is the same then we might remove "net_type" from this. If all "vmn" is the same then we might remove "fishery" and keep "net_type".

```{r}
obj$setModelParameters(qinv_names = c("species"))
obj$params_init$log_qinv

obj$setModelParameters(qinv_names = c("species", "fishery"))
obj$params_init$log_qinv

obj$setModelParameters(qinv_names = c("species", "fishery", "net_type"))
obj$params_init$log_qinv
```

9. `test_fishery_spp` - Test fishery naming convention. This allows the user to separate adult Chinook and all Chinook between the test fisheries. May also include jack Chinook at Whonnock this way.

```{r}
obj$setModelParameters(qinv_names = c("species", "fishery", "net_type"),
  test_fishery_spp = c("sockeye_whonnock", "adultchinook_whonnock", "chinook_albion"))
obj$data_info$test_fishery_spp
obj$params_init$log_qinv
```

### Set Prior/Penalty Distributions

For many of these parameters, we may not want to hold them fixed, but we will want to inform them based on prior knowledge, such as the expansion line. This can be done in a Bayesian framework by passing prior distributions to each parameter. In this case, the parameter estimates are well defined as the posterior mode. However, we may also stay in a Frequentist space and use prior information as a penalty that penalizes the likelihood to stay within the penalized region. The major difference between prior vs. penalty is simply whether or not to include the Jacobian transformation.

Setting up the penalties/priors can be done with the `setPriors`. This takes `priors` - a list of distributions that the user passes as well as `includeJacobian` - logical of whether or not to include the Jacobian. Default is to include no prior/penalties. Note that all prior distributions must be given on the log scale (`log = TRUE`).

```{r}
obj$setPriors(priors = list(sigma0 = \(x){dnorm(x, 4, 0.5, log = TRUE)},
  beta = \(x){dnorm(x, c(0,4,9), 0.5, log = TRUE)},
  qinv = \(x){dnorm(x, c(12000, 6000, 6000, 6000), 1000, log = TRUE)}),
  includeJacobian = TRUE)

obj$prior_distributions$dlog_sigma0(4) + obj$prior_jacobians$dlog_sigma0(4)
obj$prior_distributions$dbeta(c(0,4,9))
obj$prior_distributions$dlog_qinv(c(12000, 6000, 6000, 6000)) + obj$prior_jacobians$dlog_qinv(c(12000, 6000, 6000, 6000))
```

## Example 2022

Let's put it all together and have a look at what we estimate.

```{r}
speciesComp <- speciesCompModel$new(
            species = c("smallresident", "largeresident", "jackchinook", "sockeye", "adultchinook"), 
            site = "Mission", date = as.Date("2022-07-15"),  ndays = 1)

## Process all the data and set it within the R6 object.
speciesComp$processData(sonar_counts = mission_2022$sonarCounts, sonar_lengths = mission_2022$sonarLengths, 
  test_fishery_counts = list(whonnock = mission_2022$whonnock, albion = mission_2022$albion), salmon_passage_table = mission_2022$salmonPassage,
  include_tangled = TRUE, dropN = 2, feasible_lengths = c(10,120))

## Here you can change things as you play around with the inputs.
#------------------------------------------------------------------------

## Change the date and number of days to combine within the analysis.
speciesComp$setDate(date = "2022-08-15", ndays = 3)

## Change the species available.
speciesComp$setSpecies(species = c("smallresident", "largeresident", "jackchinook", "sockeye", "adultchinook"))

## Set the prior lengths, manually or from test fishery.
speciesComp$setSpeciesLengths(mu = c("smallresident" = 22, "largeresident" = 38, "jackchinook" = 42),
                       sigma = c("smallresident" = 2.5, "largeresident" = 3.5),
                        test_fishery_lengths = mission_2022$testFisheryLengths, ndays = 7)

## Set the model parameters for the model fitting:
speciesComp$setModelParameters(fixed_parameters = c("mu", "sigma"),
                                fixed_values = list(), initial_values = list(beta = c(0, 4, 9)),
                                formula_proportions = list("all" = ~ 0 + factor(day), "jackchinook" = ~ -1 + factor(day)), #
                                delta_mu_bounds = list(lower = -2.5, upper = 2.5),
                                formula_lengths = ~ 0 + SonarBin, 
                                qinv_names = c("species", "net_type"),
                                test_fishery_spp = c("sockeye_whonnock", "chinook_whonnock", "chinook_albion"))

## How many parameters are in qinv?                                
## Convenience for now, use it to set prior on qinv.
n_test <- length(speciesComp$data_info$test_fishery_input)
## Set the prior distributions to use.
speciesComp$setPriors(priors = list(N_largeresident = \(x){dnorm(x, mean = 4000, sd = 500, log = TRUE)}, 
                                    qinv = \(x){dnorm(x, mean = c(12000, 5000, 5000), sd = c(2000, 1000, 1000), log = TRUE)},
                                    beta = \(x){dnorm(x, mean = c(0, 4, 9), sd = 0.2, log = TRUE)},
                                    alpha_jackchinook = \(x){dnorm(x, mean = -2.5, sd = 0.1, log = TRUE)},
                                    delta_mu = \(x){dnorm(x, mean = 0, sd = 0.2, log = TRUE)}),
                                    includeJacobian = TRUE)

## Model fitting:
est_pars <- speciesComp$fitModel(control = list(include_test_fishery = TRUE, adjust_lengths = TRUE, test_fishery_weights = 10))
## Current avialable plots
speciesComp$plotMix(day = 1) ## Adjusted for beam spreading
speciesComp$plotMix(day = 2) ## Adjusted for beam spreading
speciesComp$plotMix(day = 3) ## Adjusted for beam spreading
speciesComp$plotBeamSpreading() ## Estimated beam spreading
speciesComp$plotTestFishery() ## Test fishery Pearson Residuals.

## Estimated Parameters:
speciesComp$params_estimated

## Total salmon estimates:
speciesComp$params_estimated$N_daily

## Proportion estimates:
speciesComp$params_estimated$p_daily
```

## Example 2025

Build the model object for the 2025 data.

```{r}
obj <- speciesCompModel$new(date = "2025-08-01", site = "Mission")

obj$processData(sonar_counts = mission_2025$sonarCounts, 
  sonar_lengths = mission_2025$sonarLengths,
  test_fishery_counts = list(whonnock = mission_2025$whonnock, albion = mission_2025$albion), 
  salmon_passage_table = mission_2025$salmonPassage, 
  include_tangled = TRUE, dropN = 2, feasible_lengths = c(10,120))
```

Now let's pick a challenging day when there are lots of sockeye and pink, such as August 20th. Note that in 2025, small resident were  not recorded.

```{r}
obj$setDate(date = "2025-08-20", ndays = 1)
obj$setSpecies(species = c("largeresident", "jackchinook", "pink", "sockeye", "adultchinook"))

## Set the prior lengths, manually or from test fishery.
obj$setSpeciesLengths(mu = c("largeresident" = 38, "jackchinook" = 42, "sockeye" = 57, "pink" = 51),
  sigma = c("largeresident" = 3.5, "pink" = 3),
  test_fishery_lengths = mission_2025$testFisheryLengths, ndays = 7)

## Set the model parameters for the model fitting:
obj$setModelParameters(fixed_parameters = c("mu", "sigma"),
  fixed_values = list(), initial_values = list(beta = c(0, 4, 9)),
  formula_proportions = list("all" = ~ 1, "jackchinook" = ~ 1), 
  delta_mu_bounds = list(lower = -2.5, upper = 2.5),
    formula_lengths = ~ 0 + SonarBin, 
    qinv_names = c("species", "net_type"),
  test_fishery_spp = c("sockeye_whonnock", "chinook_whonnock", "chinook_albion"))

obj$setPriors(priors = list(N_largeresident = \(x){dnorm(x, mean = 4000, sd = 500, log = TRUE)}, 
  qinv = \(x){c(dnorm(x[1], 12000, 1000, log = TRUE), dnorm(x[-1], mean = 5000, sd = 500, log = TRUE))},
  beta = \(x){dnorm(x, mean = c(0, 4, 9), sd = 0.1, log = TRUE)},
  alpha_jackchinook = \(x){dnorm(x, mean = -2.5, sd = 0.1, log = TRUE)},
  delta_mu = \(x){dnorm(x, mean = 0, sd = 0.5, log = TRUE)}),
  includeJacobian = TRUE)

## Model fitting:
est_pars <- obj$fitModel(control = list(include_test_fishery = TRUE, adjust_lengths = TRUE, test_fishery_weights = 100))

## Current available plots
obj$plotMix(day = 1) ## Adjusted for beam spreading
obj$plotBeamSpreading() ## Estimated beam spreading
obj$plotTestFishery() ## Test fishery Pearson Residuals.

## Mean adjustment
obj$params_estimated$delta_mu

## New Mean
obj$params_estimated$mu_adj

## Total salmon estimates:
obj$params_estimated$N_daily
## Reconstructed Sockeye Abundance Aug 20th: 178,270
## Mission Official Sockeye Abundance Aug 20th: 167,000
## Mission Official Pink Abundance Aug 20th: 23,200

## Expansion Line
obj$params_estimated$qinv

## sigma0
obj$params_estimated$sigma0
```
